---
title: "lab_09.Rmd"
author: "derek willis"
date: "2024-10-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

-   A Datawrapper account
-   Election data in CSV files in the `data` folder
-   Tidycensus

## Load libraries and establish settings

**Task** Create a codeblock and load appropriate packages and settings for this lab.

```{r}
# load libraries here
library(tidyverse)
library(tidycensus)
library(janitor)
```

## Questions

**Q1.** You're interested in seeing how the methods of voting have changed in Maryland since 2018. Using the `md_votes_by_type.csv` that has data from the top of the ticket (governor in 2018 and 2022, president in 2020), make a chart showing the percentage of votes cast in each category - early voting, election day and mail - over the three years. You'll need to make those percentages in R and then export the CSV using `write_csv`, then make the chart in Datawrapper. Have a title, description and source that tells readers about the data. The title and description should highlight the most newsworthy or important aspect.

HINT: In Datawrapper's "Check and Describe" tab, you can click on a column to get the option of "hiding" it in any visualization.

What do you think the story here is? Put the answer below along with the URL of your chart.

**A1.** The story here is definitely the changes in voting habits in 2020 especially, when the COVID-19 pandemic started. Mail-in ballot rate skyrocketed to nearly 50%. Election day ballot rate plummeted. In general, the story is how voting habits drastically changed due to the pandemic. Here is the link to my chart: https://datawrapper.dwcdn.net/wGyl6/1/

```{r}
maryland_votes <- read.csv("data/md_votes_by_type.csv")

maryland_votes <- maryland_votes |>
  rowwise() |>
  mutate(early_percent = (Early / Total) *100,
         election_percent = (Election.Day / Total) *100,
         mail_percent = (Mail / Total) *100)

write.csv(maryland_votes, "maryland_votes.csv")
```

**Q2.** You've been assigned to track early-voting turnout in Maryland at the county level. To do that, use [data from the state](https://elections.maryland.gov/press_room/2024_stats/PG24/EarlyVoting%20RAW%20data.csv). YOU DO NOT NEED TO LOAD THIS FILE. I've created several CSV files from it; each has a row for each day, with the counties as columns, so you can track early voting turnout in certain categories. The available files in the data folder are:

* ev_totals_18_24.csv - early voting turnout by the 18-24 age group
* ev_totals_65_up.csv - early voting turnout by the 65+ age group
* ev_totals_rep.csv - early voting turnout by Republicans
* ev_totals_dem.csv - early voting turnout by Democrats
* ev_totals_una.csv - early voting turnout by unaffiliated voters
* ev_totals_women.csv - early voting turnout by women
* ev_totals_men.csv - early voting turnout by men

Which file you use to make your chart, and what kind of chart is up to you (but it MUST be a chart, not a map or table), so experiment to see what works best. Since we have data over time, you might start with a line chart, but you might not end up there. You can display all counties, or just some of them - to limit the number of counties, you'll need to write code to select only the columns that you need and then use `write_csv()` to export the dataframe to a CSV file that you'll use in Datawrapper.

Your chart MUST: 

\* Have a title, description and source that tells readers about the data. The title and description should highlight the most newsworthy or important aspect. 
\* Easily readable labels and values 
\* Be published
\* Put the question you chose to answer below along, with the URL of your chart.

Tell me below why you picked the CSV you did. If you made any changes to the data, put that code below.

**A2.** I picked the 18-24 year old csv file because I feel like elections recently have been largely decided by youth turnout. We are seeing youth movements in practically every election now, so I felt it was important to analyze the data because of the relevancy. The chart showed that Prince Georges' County and Montgomery County had the highest early voting turnout among 18-24 year old people. To pick my counties, I just chose a handful of counties that I felt were the largest, most notable counties in Maryland. Here is a link to my chart: https://datawrapper.dwcdn.net/LPtLJ/1/. 

```{r}
youth_ev <- read.csv("data/ev_totals_18_24.csv")

youth_ev <- youth_ev |>
  select(c(Day, Anne.Arundel, Baltimore.City, Baltimore.County, Frederick, Howard, Montgomery, Prince.George.s))

write.csv(youth_ev, "datawrapper_youth_ev.csv")
```

**Q3.** After reading the recent CNN story about fundraising called "[How elderly dementia patients are unwittingly fueling political campaigns](https://www.cnn.com/interactive/2024/10/politics/political-fundraising-elderly-election-invs-dg/)," you decide to dig into the habits of Maryland political donors who give money via ActBlue and WinRed.

You want to show Maryland's political geography - where those donors live - and the frequency of donations. To do that, you'll load in two CSV files: `md_winred_zip_totals.csv` and `md_actblue_zip_totals.csv`.

You also will need to get population by zip code for Maryland - we did a similar thing in a previous lab, but this time you want the total population as your variable. Use tidycensus to do that, and save that dataframe, then join it to each of your contribution dataframes - you will need to make sure that the zip codes you're joining on are the same datatype.

In each dataframe, use the _number of donations_ to calculate a per capita rate of the number of donations per 1,000 population for each zip code using `mutate()`. Calculating a rate works like this: it's the thing you're counting divided by the population times the "per" number.

Then export each dataframe using `write_csv` and make two (2) choropleth maps in Datawrapper. For each map, you will need to:

-   Search for Maryland zip codes (use the 2020 data) as the base map
-   Upload the CSV file
-   Change the tooltip to show both the rate of donations and the total amount (you can put them side by side or put one under the other, as we did in the pre_lab). Make sure to properly format both numbers to make them readable.
-   Have a title, description and source that tells readers about the data (the source is the Federal Election Commission, and the data covers Jan. 1, 2023 through Oct. 16, 2024). The title and description should highlight the most newsworthy or important aspect.

Compare the two maps. What do you think are the most interesting story ideas in both, and what would you want to know more about? Put the answer below along with the URLs of your maps.

**ANSWER NUMBER 3.** 
Link to winred map: https://datawrapper.dwcdn.net/7sDT1/2/.
Link to actblue map: https://datawrapper.dwcdn.net/u7xeB/1/.

The most interesting story idea is the same in both, to me. The Chevy Chase/Bethesda area was loaded in terms of donation totals and rates. In general, metropolitan areas were heavy in this regard, but that Chevy Chase/Bethesda area (that is known to be very wealthy) had huge numbers. ActBlue also had more donations and higher donation amounts, which makes sense given the state we live in. One thing I'd like to know more about is how these donations were obtained. What efforts were put in, especially in the Chevy Chase/Bethesda area, to secure money from these rich people?

I used ChatGPT to get a framework for the code and then tweaked it a ton through trial and error until I got what worked.

ChatGPT prompt: "I have this prompt in R (give me one block of code): After reading the recent CNN story about fundraising called "[How elderly dementia patients are unwittingly fueling political campaigns](https://www.cnn.com/interactive/2024/10/politics/political-fundraising-elderly-election-invs-dg/)," you decide to dig into the habits of Maryland political donors who give money via ActBlue and WinRed.

You want to show Maryland's political geography - where those donors live - and the frequency of donations. To do that, you'll load in two CSV files: md_winred_zip_totals.csv and md_actblue_zip_totals.csv.

You also will need to get population by zip code for Maryland - we did a similar thing in a previous lab, but this time you want the total population as your variable. Use tidycensus to do that, and save that dataframe, then join it to each of your contribution dataframes - you will need to make sure that the zip codes you're joining on are the same datatype.

In each dataframe, use the _number of donations_ to calculate a per capita rate of the number of donations per 1,000 population for each zip code using mutate(). Calculating a rate works like this: it's the thing you're counting divided by the population times the "per" number."

ChatGPT answer: "Absolutely! Here's a single, complete block of R code that handles the entire process—from loading the data to calculating per capita donation rates for ActBlue and WinRed donors in Maryland by ZIP code:

r
Copy
Edit
# Load necessary libraries
library(tidyverse)
library(tidycensus)

# OPTIONAL: Set your Census API key if you haven't already (only needed once per machine)
# census_api_key("YOUR_API_KEY", install = TRUE)

# Read in ActBlue and WinRed donation data
actblue <- read_csv("md_actblue_zip_totals.csv")
winred  <- read_csv("md_winred_zip_totals.csv")

# Get total population for Maryland ZIP Code Tabulation Areas (ZCTAs)
md_pop <- get_acs(
  geography = "zip code tabulation area",
  variables = "B01003_001",  # total population
  state = "MD",
  year = 2022,
  geometry = FALSE
) %>%
  rename(zip = GEOID, total_population = estimate) %>%
  select(zip, total_population) %>%
  mutate(zip = as.character(zip))

# Ensure ZIP codes are character in donation data
actblue <- actblue %>% mutate(zip = as.character(zip))
winred  <- winred %>% mutate(zip = as.character(zip))

# Join population data and calculate donations per 1,000 people
actblue_joined <- actblue %>%
  left_join(md_pop, by = "zip") %>%
  mutate(donations_per_1000 = (donation_count / total_population) * 1000)

winred_joined <- winred %>%
  left_join(md_pop, by = "zip") %>%
  mutate(donations_per_1000 = (donation_count / total_population) * 1000)
⚠️ Notes:
Replace donation_count with the actual name of the column in your CSVs that counts the number of donations.

You can use View(actblue_joined) or glimpse(winred_joined) after running this to inspect your results.

Let me know if you want to map this or export to CSV!

Want me to help with visualizing it on a map or chart next?"


```{r}
winred_zip <- read.csv("data/md_winred_zip_totals.csv")
actblue_zip <- read.csv("data/md_actblue_zip_totals.csv")

pop_data <- get_acs(geography = "zip code tabulation area",
  variables = "B01003_001",
  year = 2020,
  survey = "acs5") |>
  mutate(GEOID = as.character(GEOID))
  
winred_zip <- winred_zip |>
  rename(GEOID = zip) |>
  mutate(GEOID = as.character(GEOID))

actblue_zip <- actblue_zip |>
  rename(GEOID = zip) |>
  mutate(GEOID = as.character(GEOID))

winred_pop <- winred_zip |>
  left_join(pop_data, by = "GEOID") |>
  mutate(donations_percapita = (donations / estimate) * 1000)

actblue_pop <- actblue_zip |>
  left_join(pop_data, by = "GEOID") |>
  mutate(donations_percapita = (donations / estimate) * 1000)

write.csv(winred_pop, "datawrapper_winred.csv")
write.csv(actblue_pop, "datawrapper_actblue.csv")
```
