---
title: "Test run"
output: html_document
date: "2025-04-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(janitor)
```

```{r}
independent_expenditures <- read.csv("independent_expenditure_2024.csv")
```

```{r}
independent_expenditures <- independent_expenditures |>
  clean_names() |>
  rename(
    candidate_id = cand_id,
    candidate_name = cand_name,
    spender_id = spe_id,
    spender_name = spe_nam,
    election_type = ele_type,
    candidate_state = can_office_state,
    candidate_district = can_office_dis,
    candidate_office = can_office,
    candidate_party = cand_pty_aff,
    expenditure_amount = exp_amo,
    expenditure_date = exp_date,
    aggregate_amount = agg_amo,
    support_or_oppose = sup_opp,
    purpose = pur,
    payee = pay,
    file_number = file_num,
    amendment_indicator = amndt_ind,
    transaction_id = tran_id,
    image_number = image_num,
    receipt_date = receipt_dat,
    election_year = fec_election_yr,
    previous_file_number = prev_file_num,
    dissemination_date = dissem_dt
  )
```

##importing after cleaning in OpenRefine

```{r}
cleaned_expenditures <- read.csv("updated_cleaned.csv")
```

There are 23 columns and 72,971 rows. This data represents all of the independent expenditure spending by candidates/committees from 2023-2024. The sheer amount of data might be limiting because there could be errors in the data. It also accounts for all types of races, so to compare certain types of expenditures would require a lot of filtering. Our questions might be hard to answer in terms of finding average spending because of the amount of data. Additionally, comparing geographical spending compared to other years will be hard because the dataset lacks geographical information.

Question 1: Which presidential candidates have spent the most during their campaign? Who has spent the least?

ANSWER: Jon Tester has the most, eight candidates are tied for the least.

```{r}
cleaned_expenditures |>
  group_by(candidate_name) |>
  summarise(total = sum(expenditure_amount, na.rm = TRUE)) |>
  arrange(desc(total))
```

Question 2: What are the most common payment purposes for payment in federal campaign finance data? What is the average amount paid per transaction for these top purposes?

ANSWER: 

Phonebanking -> average amount $980.15
Production Costs -> average amount $177,963.6
Media Production -> average amount $9263.04
Advertisement Management Services -> average amount $928.37
Canvassing -> average amount $65,250.19

```{r}
top_purposes <- cleaned_expenditures |>
  filter(!is.na(purpose), purpose != "") |>
  group_by(purpose) |>
  summarise(
    count = n(),
    average_amount = mean(expenditure_amount, na.rm = TRUE),
    total_amount = sum(expenditure_amount, na.rm = TRUE)
  ) |>
  arrange(desc(count)) |>
  slice_head(n = 10)

top_purposes
```

Question 3: Which candidates receive the most independent support versus opposition, and how does spending differ between supporting and attacking candidates?

ANSWER: Jon Tester received the most independent support versus opposition. Democrats tend to spend more money on elections and their campaigns compared to Republicans.

```{r}
summary <- cleaned_expenditures |>
  filter(!is.na(candidate_name), support_or_oppose %in% c("S", "O")) |>
  group_by(candidate_name, support_or_oppose) |>
  summarise(total_spent = sum(expenditure_amount, na.rm = TRUE), .groups = "drop")

top_supported <- summary |>
  filter(support_or_oppose == "S")|>
  arrange(desc(total_spent)) |>
  head(5)

top_opposed <- summary |>
  filter(support_or_oppose == "O") |>
  arrange(desc(total_spent)) |>
  head(5)

purpose_comparison <- cleaned_expenditures |>
  filter(!is.na(purpose), support_or_oppose %in% c("S", "O")) |>
  group_by(candidate_name, support_or_oppose) |>
  summarise(total_spent = sum(expenditure_amount, na.rm = TRUE), .groups = "drop") |>
  arrange(desc(total_spent))

purpose_comparison
```

Question 4: Which PAC's donated in both the 2020 and 2024 election?

ANSWER: PAC's that donated in both elections are:

- American Crossroads				
- Americans for Constitutional Liberty dba The Conservative Caucus				
- End Citizens United				
- Gun Rights America				
- Indivisible Action				
- MONTANA HUNTERS AND ANGLERS LEADERSHIP FUND				
- MORE JOBS, LESS GOVERNMENT				
- March On PAC				
- Montana Rural Voters (WORC)				
- NEA Advocacy Fund

```{r}
cleaned_expenditures_2020 <- read.csv("cleaned_2020_expenditures.csv")
```

```{r}
cleaned_expenditures_2020 <- cleaned_expenditures_2020 |>
  clean_names() |>
  rename(
    candidate_id = cand_id,
    candidate_name = cand_name,
    spender_id = spe_id,
    spender_name = spe_nam,
    election_type = ele_type,
    candidate_state = can_office_state,
    candidate_district = can_office_dis,
    candidate_office = can_office,
    candidate_party = cand_pty_aff,
    expenditure_amount = exp_amo,
    expenditure_date = exp_date,
    aggregate_amount = agg_amo,
    support_or_oppose = sup_opp,
    purpose = pur,
    payee = pay,
    file_number = file_num,
    amendment_indicator = amndt_ind,
    transaction_id = tran_id,
    image_number = image_num,
    receipt_date = receipt_dat,
    election_year = fec_election_yr,
    previous_file_number = prev_file_num,
    dissemination_date = dissem_dt
  )
```

```{r}
spenders_2020 <- cleaned_expenditures_2020 |>
  select(spender_name) |>
  distinct()

spenders_2024 <- cleaned_expenditures |>
  select(spender_name) |>
  distinct()

common_spenders <- inner_join(spenders_2020, spenders_2024, by = "spender_name")

common_spenders
```

Question 5: How are Republicans and Democrats spending their money differently? What are each party’s top spending points?

ANSWER: Democrats spent a lot of money on television ads, radio and media production. Republicans spent a lot of money on data harvesting. The Democrats top spending point was PLACED MEDIA: TV. The Republicans top spending point was Data Harvesting.

```{r}
cleaned_expenditures |>
  group_by(candidate_party, expenditure_amount, purpose) |>
  summarise(total = sum(expenditure_amount, na.rm = TRUE)) |>
  arrange(desc(total))
```

