---
title: "lab_02"
author: "derek willis"
date: "2024-08-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
library(lubridate)
```

## Load Data

For this lab we'll be working with a slice of Maryland voter registration data: records of people who have registration dates this year through Aug. 7. Let's load it.

```{r}
md_voters_2024 <- read_rds("data/lab_02.rds")
```

## Examine Data

Take a look at the data, getting familiar with what it contains. Most of the columns should be self-explanatory, but there are two date columns: STATE_REGISTRATION_DATE and COUNTY_REGISTRATION_DATE. The former represents the date the person registered as a Maryland voter; the latter is when they registered in their current county. They may be the same, but if a voter moves from one county to another they will differ. This is important because if the two dates are the same, that's a pretty good indication that this is a *new* voter, someone who has no history of registering in Maryland. You'll need to keep that definition in mind.

## Answer questions

**Q1** How many voters appear to be *new* registrants since July 21, when President Joe Biden announced he would leave the race, based on the information above? Create a dataframe to hold the only those voters. Which counties have seen the most? Take a look at county population figures here and compare the two in your answer: https://worldpopulationreview.com/us-counties/maryland.

**A1** Montgomery County, Baltimore City and Baltimore County have seen the most, according to the data. This matches up pretty well with county population figures, where Montgomery County is #1 as well. Baltimore County and Baltimore City are #3 and #5 respectively, but it's within the same ballpark. Prince George's County is #2 on the county population list while being #4 on this dataset.

Also -- I used ChatGPT a bit to figure out the as.Date command. The prompt I gave was along the lines of "What function do I use in the R tidyverse to filter a set of dates from a dataset to only consider the occurrences after a specific date." Just a disclaimer.

```{r}
md_voters_2024 |>
  filter(STATE_REGISTRATION_DATE == COUNTY_REGISTRATION_DATE, STATE_REGISTRATION_DATE > as.Date("2024-07-21")) |>
  mutate(NEW_VOTERS ="Yes") |>
  group_by (COUNTY) |>
  summarise(NEW_VOTERS = n()) |>
  arrange(desc(NEW_VOTERS))
```

**Q2** A better question is whether the new voters you found in Q1 are an outlier compared to other weeks this year. Using the original dataframe with 2024 registrants, use `mutate()` and lubridate to produce week numbers from the COUNTY_REGISTRATION_DATE (Lubridate has a function that turns a date into a week number. Use that to answer this question). Save that to the original dataframe. Then calculate how many voters are in each week. Describe the distribution of voters groups in a sentence - are the weeks from July 21-August 7 different from the rest, and if so, how?

**A2** Yes, those weeks are a bit different than the others. Both of those weeks clear 7000 new voters while most other weeks fall well short of that mark. This would indicate higher voter registration in those two weeks.

Also, I spent like an hour and a half on this question alone. Not sure if it's right. I think it is. Maybe.

```{r}
md_voters_2024 |>
  mutate(week = floor_date(COUNTY_REGISTRATION_DATE, "week")) |>
  group_by (week) |>
  summarise(NEW_VOTERS = n()) |>
  arrange(week)
```

**Q3** Using lubridate, show which month saw the largest number of registrations. Does the answer help shed more light on Q2?
**A3** July. And yes, it checks out. July (and those weeks) is definitely different, and higher, compared to the rest.

```{r}
md_voters_2024 |>
  mutate(month = floor_date(COUNTY_REGISTRATION_DATE, "month")) |>
  group_by (month) |>
  summarise(NEW_VOTERS = n()) |>
  arrange(month)
```

**Q4** Maryland has, officially, five regions: https://www.visitmaryland.org/article/maryland-regions. Modify the original dataframe again to add a `region` column based on the `COUNTY` value using case_when. Are you going to have to do this for every Maryland county plus Baltimore City? Yes, you are. Which regions and parties account for the most new voters?

**A4** The Capital Region and Central Maryland account for the most new voters by far. As for parties, DEM and UNA account for the most. These regions and parties dominate the top of the new voters list. Another disclaimer, I used ChatGPT to figure out the case_when section. I asked how to mutate in R tidyverse using the case_when function and a region. After messing around with it for about an hour and trying a ton of different things, I finally got this to work.

```{r}
md_voters_2024 |>
  filter(COUNTY_REGISTRATION_DATE == STATE_REGISTRATION_DATE) |>
  mutate(region = case_when(
    COUNTY %in% c("Garrett", "Allegany", "Washington") ~ "Western Maryland",
    COUNTY %in% c("Montgomery", "Frederick", "Prince George's") ~ "Capital Region",
    COUNTY %in% c("Baltimore City", "Baltimore County", "Carroll", "Harford", "Howard", "Anne Arundel") ~ "Central Maryland",
    COUNTY %in% c("Charles", "Calvert", "Saint Mary's") ~ "Southern Maryland",
    COUNTY %in% c("Cecil", "Kent", "Queen Anne's", "Talbot", "Caroline", "Dorchester", "Wicomico", "Worcester", "Somerset") ~ "Eastern Shore",
  )) |>
  group_by(region, COUNTY, PARTY) |>
  summarise(NEW_VOTERS = n()) |>
  arrange(desc(NEW_VOTERS))

```

**Q5** What's an additional question you could ask of this data?

**A5** You could go even deeper. Which congressional districts account for the most new voters? There is a CONGRESSIONAL_DISTRICTS column that can be used for this -- and it could filter even more specific data than what we already have.
