---
title: "lab_06"
author: "derek willis"
date: "2024-10-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

-   Tabula

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse and any other libraries you'll need
library(tidyverse)
```

## Get Our PDF

We'll be working with the [Maryland monthly voter registration update](https://elections.maryland.gov/pdf/vrar/2024_09.pdf) from the State Board of Elections. You'll want to download it to a place you'll remember (like your Downloads folder, or the labs folder in your repository). The goal is to write a couple of paragraphs that summarize the changes in voter registration in Maryland during the month of September.

## Setup Tabula

Start Tabula, then go to <http://127.0.0.1:8080/> in your browser. Click the "Browse" button and find the PDF file and click "open", and then click the "Import button" in Tabula. This will take a few seconds.

This PDF has multiple possible tables to extract. We're going to make four dataframes: 

* new registrations by party
* removals by party
* changes in registration
* current registrations by county.

You will need to draw boxes around four sections and export four CSV files that you will load into R. In Tabula, draw a box around the borders of those tables and click the "Preview & Export Extracted Data" button for each one. The export should look pretty clean, but might be missing some column headers. You'll need to include total columns.

Export each CSV (each one should be called `tabula-2024-09.csv` by default, so you should rename them for clarity) to your lab_06/data folder.

From there, you will need to read in and clean up the data, and in particular the headers. You can choose to include the headers from the PDF in your exported CSV files OR to exclude them and add them when importing. The pre_lab has an example of how to do this!

## Load and clean up the data in R

You will need to read in and clean up the data so that it can be used for analysis. By "clean" I mean the column headers must not contain spaces and they should have meaningful names, not "x1" or something similar. The current registrations dataframe *MUST* include the county name, but should not include the columns under `Changes` and `Party Affiliation From`. How you do that is up to you (and could involve something outside R), but you can use `select()` with or without the minus sign to include or exclude certain columns. You also can use the `rename` function to, well, rename columns.

```{r}
# Party registrations
party_registrations <- read_csv("tabula-new-registration.csv")
party_registrations
```

```{r}
# Removals
voter_removals <- read_csv("tabula-removals.csv")
voter_removals
```

```{r}
# Changes
voter_changes <- read_csv("tabula-changes.csv")
voter_changes
```

```{r}
# Current registrations
current_registrations <- read_csv("tabula-current-registrations.csv")
current_registrations
```

## Rinse, Repeat: September 2022 data

Do all of that again, for the [September 2022 voter registration update](https://elections.maryland.gov/pdf/vrar/2022_09.pdf), so we can compare the two. If you've done well, this is easy: just do the same steps in Tabula, export the files and use the same code to read in and clean up the data, *just remember to give these dataframes different names* so that you can compare September 2024 to September 2022.

To answer the following questions, you'll need to calculate percentages and do other calculations. Whether you do that by combining or joining dataframes or you write code to run separately on both is up to you.

```{r}
# Party registrations
party_registrations_2022 <- read_csv("tabula-new-registration-2022.csv")
party_registrations_2022
```

```{r}
# Removals
voter_removals_2022 <- read_csv("tabula-removals-2022.csv")
voter_removals_2022
```

```{r}
# Changes
voter_changes_2022 <- read_csv("tabula-changes-2022.csv")
voter_changes_2022
```

```{r}
# Current registrations
current_registrations_2022 <- read_csv("tabula-current-registrations-2022.csv")
current_registrations_2022
```

## Answer questions

Q1. Calculate the percentage of registered Democrats, Republicans and unaffiliated voters for the entire state in September 2024 and September 2022, and then calculate the percentage change for those three party categories (the formula for percentage change is easy to remember: (New-Old)/Old). How would you describe the changes in these partisan categories in a story? What's the most interesting thing about these changes that would be the lede of that story?

A1. The Democrats saw close to a 3% decrease in number of registered voters. Republicans saw a decrease of about half a percentage point. Meanwhile, the unaffiliated voter category saw over an 8% increase --- which would probably be the story and lede here. The story overall would be something about how unaffiliated registration numbers have seen a substantial increase and while both major parties have seen decreases in their registration, the Democrats have been hit the hardest.

There was a slight bit of ChatGPT used here. I had some original code (basically what you see here, but without the suffix part) that wasn't working, so I plugged it into the AI and asked for a fix and they provided me the suffix command. Everything worked after that.


```{r}
library(dplyr)

data_2022 <- data_frame(
  party = c("DEM", "REP", "UNAF"),
  total_registered = c(2250961, 998905, 841573)
)

data_2024 <- data_frame(
  party = c("DEM", "REP", "UNAF"),
  total_registered = c(2219417, 1009323, 923927)
)

combined_data <- data_2022 |>
  left_join(data_2024, by = "party", suffix = c("_2022", "_2024")) |>
  mutate(
    percentage_2022 = total_registered_2022 / sum(total_registered_2022) * 100,
    percentage_2024 = total_registered_2024 / sum(total_registered_2024) * 100,
    percentage_change = (percentage_2024 - percentage_2022) / percentage_2022 * 100
  )

combined_data
```

Q2. Among the Maryland jurisdictions with the most registered voters, which one had the biggest percentage increase in unaffiliated voters between 2022 and 2024? How did that county vote for governor in 2022 and for president in 2020? Contrast the partisan percentages of that county with the county that had the eighth-largest percentage increase in unaffiliated voters. You should display information from the data that shows the counties with the largest percentage increase first.

A2. The jurisdiction with the biggest percentage increase in unaffiliated voters was Baltimore County with a near 84% increase from 2022 to 2024. The county voted for Joe Biden for president in 2020, and Wes Moore for governor in 2022. The lowest percentage change was Baltimore City, with a little over 11% decrease from 2022 to 2024. 

```{r}
library(dplyr)

voter_data <- data.frame(
  jurisdiction = c("MONTGOMERY", "PR. GEORGE'S", "ANNE ARUNDEL", "HOWARD", "BALTIMORE CO.", "BALTIMORE CITY", "HARFORD", "FREDERICK"),
  total_registered_2022 = c(160577, 86155, 98350, 57367, 108672, 63933, 45260, 54268),
  total_registered_2024 = c(173554, 100970, 105506, 63155, 199734, 56663, 41705, 49092)
)

voter_data |>
  mutate(percentage_change = (total_registered_2024 - total_registered_2022) / total_registered_2022 * 100) |>
  arrange(desc(percentage_change))
```

Q3. Do the party percentages of new registrants (% Dem, % Rep, % Unaffiliated) from September 2024 and September 2022 match the overall percentage of partisan registration in the state at those times?

In other words, did new registrants mirror the overall partisan composition of the electorate, or were they different? Do any of the methods of registration vary significantly from the overall partisan pattern? Which method is worthy of further exploration?

A3. Nope. They were different. Republicans saw close to a 22% increase in 2024, despite being a small portion (15%) of partisan registration in 2022. So I can't say it matched. This is probably worthy of further explanation.

Similar to question 1, I used ChatGPT a little bit for this one. I had original code similar to the finished product that wasn't working, so I plugged it into the AI and fiddled around for a fix. After a few attempts, it worked.

```{r}
library(dplyr)

new_registrants_2022 <- data_frame(party = c("DEM", "REP", "UNAF"), 
  total_registered = c(19428, 5661, 11513))

new_registrants_2024 <- data_frame(party = c("DEM", "REP", "UNAF"), 
  total_registered = c(19077, 6977, 10975))

combined_data <- merge(new_registrants_2022, new_registrants_2024, by = "party") |>
  mutate(
    percentage_2022 = total_registered.x / sum(total_registered.x) * 100,
    percentage_2024 = total_registered.y / sum(total_registered.y) * 100,
    percentage_change = ((percentage_2024 - percentage_2022) / percentage_2022) * 100
  )

combined_data
```

Q4. Choose your own adventure! Using both the 2024 and 2022 data for one of the dataframes (changes, removals or new registrations), ask and answer a question that makes at least one comparison and requires code to answer. That could involve calculating a percentage, percentage change or the raw difference between two columns.

A4. The actual code I wrote is somewhat similar to question 2, but it answers a vastly different question: how did Republican registrations change from 2022 to 2024 in Maryland's eight largest jurisdictions? You'll notice from the code that in six of the eight jurisdictions, the change was positive for Republicans. In Prince George's County especially, there was almost a 6% increase in total Republican registrations from 2022 to 2024. We know that not only did Trump win the election, but he made gains in the vast, vast majority of jurisdictions across the country from his 2020 to 2024 campaigns. The data I wrote shows those gains and where they were most prevalent. If Republican registration goes up in a jurisdiction, it is safe to assume votes for Trump did too, right? This could easily be turned into a data story showcasing which counties were most directly responsible for the "red wave" that occurred, and could help us analyze the election more closely. 

```{r}
library(dplyr)

voter_data <- data.frame(
  jurisdiction = c("MONTGOMERY", "PR. GEORGE'S", "ANNE ARUNDEL", "HOWARD", "BALTIMORE CO.", "BALTIMORE CITY", "HARFORD", "FREDERICK"),
  total_republican_registrations_2022 = c(98170, 37836, 130762, 49372, 138270, 28218, 80695, 68809),
  total_republican_registrations_2024 = c(98743, 39974, 129562, 48580, 139338, 29102, 81530, 68925)
)

voter_data |>
  mutate(percentage_change = (total_republican_registrations_2024 - total_republican_registrations_2022) / total_republican_registrations_2022 * 100) |>
  arrange(desc(percentage_change))
```

Q5. What's the best story idea or question you've seen as a result of the work you've done in this lab?

A5. I feel like the obvious one here is unaffiliated voter changes. It would appear unaffiliated voters have skyrocketed recently, with an 8% statewide increase. As noted in question 2, Baltimore County saw a whopping 84% increase in unaffiliated voters. That's an insane number, and numbers like that probably reflect the electorates discontent with the two party system (as well as the unappealing natures of most modern day Democrat and Republican candidates). There is absolutely a story here as to why the unaffiliated voter numbers are increasing. I have my thoughts as to why, but there is certainly a story to be researched and told.
